\chapter{Complexity of a-transducers}
\label{chap:complexity}

\paragraph{}
This section is concerned with the complexity of a-transducers. Since the majority of the results published to this date involve sequential transducers and sequential functions, we try to investigate two new concepts in this area - nondeterminism and the fact, that we deal with pairs of languages, without exactly defined transduction.

\paragraph{}
However, at this time we do not have any universal way of proving the minimality of an a-transducer (with respect to the number of states). For this reason, we look at some special classes of transformations and languages and present the results concerning these. This allows to present the basic concepts and opens a possibility to initiate similar study for other types of languages.

\paragraph{}
It is easy to see that for a regular language $R$ there always exists an a-transducer with $\C{R}$ states, which generates $R$ "from scratch", regardless of the input. It suffices to take the minimal finite automaton for $R$ and alter its transition function from reading to generating symbols.

\paragraph{}
Formally, for an automaton $A = (K, \Sigma, \delta, q_0, F)$ we can construct an a-transducer $M = (K, \Sigma, \Sigma, H, q_0, F)$, where $H = \{ (p, \varepsilon, a, q) | \delta (p, a) = q\} \cup \{ (q_F, a, \varepsilon, q_F) | a \in \Sigma, q_F \in F \} $. We can look at the computation of $A$ as a sequence of pairs $(q, a)$, where in each step, $A$ is in the state $q$ and reads symbol $a$. The a-transducer $M$ will work in the same way, except instead of reading symbol $a$, $M$ reads in each step $\varepsilon$ and writes $a$ on the output. Then, in the final state, $M$ consumes the whole input without generating any output. It is easy to see, that for any nonempty language $L$, $M(L) = R$.


%----------------------------------------------------------------------------
\section{"Modular-Counting" Languages}

\paragraph{}
By "modular counting" languages we understand languages in the form \\
\centerline{$L_k = \{ a^k | k \equiv 0 \pmod{k} \} $.}

\paragraph{}
We would now like to present our results concerning the minimum complexity of an a-transducer for a pair of modular counting languages.

\paragraph{}
\oznacenie By $\gcd(k,l)$ we denote a greatest common divisor of integers $k,l$, by $\lcm(k,l)$ their lowest common multiple.

\paragraph{}
\clema For a pair of languages $L_k, L_l$, the minimal state complexity of an a-transudcer $M$, such that $M(L_k) = L_l$, is 
\begin{enumerate}
\item $l$, if $k$ and $l$ are coprime integers,
\item $\frac{l}{\gcd(k,l)}$, if $k \leq l$,
\item $\min(l,\frac{k}{\gcd(k,l)})$, if $l < k < l^2$,
\item $l$, if $k \geq l^2$.
\end{enumerate}

\paragraph{}
\dokaz For the sake of clarity, we prove the four parts of the Lemma separately. However, as stated before, $l$ states are always sufficient, so we have a natural upper bound for parts 1. and 4.
\begin{enumerate}
\item Let $M = (K, \{a\}, \{a\}, H, q_0, F)$ be an a-transducer, such that $M(L_k) = L_l$. Let $M$ have $l-1$ states. Now, let us look at an accepting computation (in this case the corresponding sequence of states) of $M$ on some sufficiently long word $x \in L_k$ ($|x| \geq l$), on which $M$ generates a word $y \in L_l$. Clearly, there has to be a cycle, i. e. the computation has a form $q_0, q_1, ..., q_i, ..., q_j, ..., q_F$, where $q_F \in F$ and $q_i = q_j$, where $j < i + l$ (we assume that this is the shortest cycle in the computation, during which $M$ generates a non-empty output). In this cycle, $M$ reads a subword $a^r$ and generates an output $a^s$ for some $r,s; 1 \leq r,s \leq l-1$.

\paragraph{}
Now, let us take two longer inputs $x' = x.a^{k.r}$ and $x'' = x.a^{2k.r}$. On these two inputs, $M$ generates outputs $y' = y.a^{k.s}$ and $y'' = y.a^{2k.s}$, respectively. Since $k$ and $l$ are coprime integers and $s < l$, $k.s$ is not divisible by $l$ (the least common multiple of two coprimes is their product), therefore at least one of these outputs does not belong to $L_l$, while both $x', x'' \in L_k$. We have generated an incorrect output, thus $M$ cannot have less than $l$ states.

\item Since the case $\gcd(k,l)=1$ was treated in 1, we can assume $\gcd(k,l) > 1$. Therefore, in what follows we assume, that $\frac{l}{\gcd(k,l)} < l$.

\paragraph{}
First we will show, that $\frac{l}{\gcd(k,l)}$ states suffice. We can construct an a-transducer $M = (K, \{ a\}, \{ a\}, H, q_0, F)$, where
\begin{itemize}
\item $K = \{ q_0, q_1,  ..., q_{\frac{l}{\gcd(k,l)}-1 }\}$
\item $F = q_0$%q_{\frac{l}{\gcd(k,l)}-1 }$
\item $H = \{(q_i, a, a, q_{i+1})| 0 \leq i < \frac{k}{\gcd(k,l)}-1 \} \cup \{(q_i, \varepsilon, a, q_{i+1})| \frac{k}{\gcd(k,l)}-1 \leq i < \frac{l}{\gcd(k,l)}-2 \} \cup \{ (q_{\frac{l}{\gcd(k,l)}-1}, \varepsilon, a, q_0) \}$.
\end{itemize}

\paragraph{}
It is easy to see, that the number of iterations of this cycle on a correct input (from $L_k$) is divisible by $\gcd(k,l)$. Each iteration creates $\frac{l}{\gcd(k,l)}$ symbols $a$ on the output, therefore $M(L_k) = L_l$.

\paragraph{}
Now we need to prove, that this number really forms a lower bound for the state count. Suppose, that there is an a-transducer $M' = (K, \{a\}, \{a\}, H, q_0, F)$ with at most $\frac{l}{\gcd(k,l)} -1$ states, such that $M'(L_k) = L_l$. Similarly to the proof of part 1., we look for a cycle, in this case of the length of at most $\frac{l}{\gcd(k,l)} - 1$ states. With very similar series of arguments, we can construct two inputs $x' = x.a^{k.r}$ and $x'' = x.a^{2k.r}$, which produce outputs $y' = y.a^{k.s}$ and $y'' = y.a^{2k.s}$, respectively. If both $|y'|, |y''|$ were divisible by $l$, then also $k.s$ would be divisible by $l$. However, this is not possible, since $s < \frac{l}{\gcd(k,l)}$ and as we know from the number theory, $\lcm(k,l) = \frac{k.l}{\gcd(k,l)}$.  

\item Just like in part 2., we show, that if $k > l \land k < l^2$, then $\frac{k}{\gcd(k,l)}$ states is enough. The corresponding a-transducer will look as follows: $M = (K, \{ a\}, \{ a\}, H, q_0, F)$, where 

\begin{itemize}
\item $K = \{ q_0, q_1,  ..., q_{\frac{k}{\gcd(k,l)}-1 }\}$
\item $F = q_0$
\item $H = \{(q_i, a, a, q_{i+1})| 0 \leq i < \frac{l}{\gcd(k,l)}-1 \} \cup \{(q_i, a, \varepsilon, q_{i+1})| \frac{l}{\gcd(k,l)}-1 \leq i < \frac{k}{\gcd(k,l)}-2 \} \cup \{ (q_{\frac{k}{\gcd(k,l)}-1}, \varepsilon, a, q_0) \}$.
\end{itemize}
\paragraph{}
For similar reason as in part 2., it is clear, that $M(L_k) = L_l$.

\paragraph{}
However, the second part of the proof is a little bit different. We will not show, that an a-trandsucer $M' = (K', \{a\}, \{a\}, H', q'_0, F')$ with fewer states, such that $M'(L_k) = L_l$, generates an incorrect output, but we claim, that it is not able to generate all correct outputs (i. e., all words from language $L_l$). Let us consider the shortest nonempty  word, that we can generate from $L_k$ using $M'$?

\paragraph{}
We have assumed, that $k < l^2$, therefore we can also state, that $\frac{k}{\gcd(k,l)} < l$. Once again, we look for a cycle in the computation of $M'$. Since $|Q'| < l$, to produce an output of length $l$ the computation must have a form $q'_0, q'_1, ...,$ $ q'_i, ..., q'_j, ..., q'_F$, where $q'_F \in F'$ and $q'_i = q'_j$ for some $i$ and $j$, where $j < i + \frac{k}{\gcd(k,l)}$. In each iteration of this cycle, $M'$ has to output at least one symbol $a$.

\paragraph{}
We claim, that in each iteration of the cycle (i. e. in any of all possible cycles in its computation), $M'$ has to generate at least $\frac{l}{\gcd(k,l)}$ symbols $a$. Really, in the proof of the second part of our Lemma we have seen, that if the number $s$ - the number of output symbols generated in one iteration of the cycle - is smaller than $\frac{l}{gcd(k,l)}$, $M'(L_k) \cap L_l^c \neq \emptyset$, which leads to a contradiction.

\paragraph{}
Moreover, since $|Q'| < \frac{k}{\gcd(k,l)}$, we also know, that in one iteration of each cycle, $M'$ reads less than $\frac{k}{\gcd(k,l)}$ symbols. Now, the shortest nonempty word from $L_k$ (if $M'(\varepsilon) \neq \emptyset$, it could be trivially proven, that $M'(L_k)$ contains also words not from $L_l$) is $a^k$. The total number of iterations of all cycles is hence more than $\frac{k}{\frac{k}{\gcd(k,l)}} = \gcd(k,l)$. However, as we have claimed, every cycle generates at least $\frac{l}{\gcd(k,l)}$ symbols. Then, the smallest output length $n > \gcd(k,l).\frac{l}{\gcd(k,l)} = l$, hence we have no way to generate the word $a^l \in L_l$.

\item The correctness of the lower bound $l$ is clear from the construction based on its final automaton (see above). The impossibility of existence of a smaller a-transducer follows directly from previous part of Lemma - if $k \geq l^2$, then $l \leq \frac{k}{\gcd(k,l)}$.

\end{enumerate} \qed

\paragraph{}
As a direct consequence of Lemma 12 we obtain the following theorem.

\paragraph{}
\cveta $\C{L_k, L_l} = \min (l, \frac{\max (k,l)}{\gcd (k,l)})$.

